
```{r, eval = FALSE}
install.packages("ape")
```

```{r, eval = FALSE}
library(ape)
```

```{r}
primatedata <- read.table("Primatedata.txt", sep = "\t", header = TRUE)
```

```{r}
str(primatedata)
```

```{r}
head(primatedata)
```

```{r}
names(primatedata)
```
This gives you the names of the columns.

```{r, eval = FALSE}
primatedata
```

```{r}
primatetree <- read.nexus("consensusTree_10kTrees_Version2.nex") 
```

Let’s examine the tree by typing:

```{r}
primatetree
str(primatetree)
```

```{r , eval = FALSE}
plot(primatetree)
```

```{r, eval = FALSE}
plot(primatetree, cex = 0.5)
```

```{r, eval = FALSE}
zoom(primatetree, list(grep("Cercopithecus", primatetree$tip.label)), 
     subtree = FALSE)
```

```{r, eval = FALSE}
zoom(primatetree, list(grep("Cercopithecus", primatetree$tip.label)), 
     subtree = TRUE)
```

```{r}
primatetree2 <-drop.tip(primatetree, "Aotus_azarae_infulatus")
str(primatetree2)
```

```{r, eval = FALSE}
?plot.phylo
```

```{r, eval = FALSE}
par(mfrow = c(1,1)) 
plot(primatetree, type = "fan", edge.color = "deeppink", 
     tip.color = "green", cex = 0.5)
```

```{r, eval = FALSE}
plot(primatetree)
axisPhylo()
```

```{r}
is.binary.tree(primatetree) # we want this to be TRUE
primatetree <- multi2di(primatetree) 
```

```{r, eval = FALSE}
primatetree.reroot <- root(primatetree, "Saimiri_sciureus")
plot(primatetree.reroot)
```
```{r}
primatedata$Binomial <- gsub(" ", "_", primatedata$Binomial)
```

```{r}
row.names(primatedata) <- primatedata$Binomial

```{r}
par(mfrow = c(2, 2))
```

```{r, eval = FALSE}
hist(primatedata$AdultBodyMass_g)
hist(primatedata$GestationLen_d)
```

```{r, eval = FALSE}
hist(log(primatedata$AdultBodyMass_g))
hist(log(primatedata$GestationLen_d))
```

```{r}
hist(log(primatedata$GestationLen_d), col = rainbow(8))
```

```{r}
par(mfrow = c(1, 1))
```

```{r}
model.ols <- lm(log(GestationLen_d) ~ log(AdultBodyMass_g), 
                data = primatedata)
summary(model.ols)
```


```{r, eval = FALSE}
plot(log(GestationLen_d) ~ log(AdultBodyMass_g), data = primatedata)
abline(model.ols)
```

```{r, eval = FALSE}
points(log(GestationLen_d[Family == "Cercopithecidae"]) ~ 
       log(AdultBodyMass_g[Family == "Cercopithecidae"]), 
       data = primatedata, col= "blue",  pch = 16)
```

```{r}
primate <- comparative.data(phy = primatetree, data = primatedata, 
                            names.col = Binomial, vcv = TRUE, 
                            na.omit = FALSE, warn.dropped = TRUE)
```
```{r}
primate$dropped$tips
primate$dropped$unmatched.rows
```

```{r}
model.pgls <- pgls(log(GestationLen_d) ~ log(AdultBodyMass_g), 
                   data = primate, lambda='ML')
summary(model.pgls)
```

```{r, eval = FALSE}
plot(log(GestationLen_d) ~ log(AdultBodyMass_g), data = primate$data)
abline(model.pgls)
```

```{r}
model.pgls2 <- pgls(log(GestationLen_d) ~ log(AdultBodyMass_g), 
                    data = primate, lambda='ML', 
                    bounds = list(lambda = c(1e-6, 1)))
summary(model.pgls2)
```

```{r, eval = FALSE}
lambda.profile <- pgls.profile(model.pgls, 'lambda')
plot(lambda.profile)
```

```{r}
pgls.confint(model.pgls, 'lambda')
```

```{r}
est.lambda <- pgls(log(GestationLen_d) ~ 1, data = primate, lambda = 'ML')
summary(est.lambda)
```

```{r, eval = FALSE}
par(mfrow = c(2, 2))
plot(model.pgls)
```

```{r}
pgls.residuals <- residuals(model.pgls, phylo = TRUE)
```

```{r}
std.residuals <- pgls.residuals/sqrt(var(pgls.residuals))[1]
```

```{r}
rownames(std.residuals) <- rownames(model.pgls$residuals)
```

```{r}
rownames(std.residuals)[(abs(std.residuals) > 3)]
```

```{r}
primate.no.outliers <- primate[-which(abs(std.residuals) > 3), ]
```

```{r}
model.pgls.no.outliers <- pgls(log(GestationLen_d) ~ log(AdultBodyMass_g), 
                               data = primate.no.outliers, lambda='ML')
summary(model.pgls.no.outliers)
```

```{r, eval = FALSE}
par(mfrow = c(2, 2))
plot(model.pgls.no.outliers)
```




```{r}
lngest <- log(primatedata$GestationLen_d)
names(lngest) <- primatedata$Binomial
```

```{r}
Kcalc(lngest[primatetree$tip.label], primatetree)
```

```{r}
phylosignal(lngest[primatetree$tip.label], primatetree, reps = 1000)
```

```{r}
setdiff(primatetree$tip.label, primatedata$Binomial)
```

```{r}
primatetree2 <-drop.tip(primatetree, setdiff(primatetree$tip.label, 
                        primatedata$Binomial))
```

```{r}
setdiff(primatedata$Binomial, primatetree$tip.label)
```

```{r}
matches <- match(primatedata$Binomial, primatetree2$tip.label, nomatch = 0)
primatedata2 <-subset(primatedata, matches != 0)
```

```{r}
str(primatedata2)
str(primatetree2)
```
