---
title: "Visualising phylogenies in R"
author: "Natalie Cooper (natalie.cooper@nhm.ac.uk)"
date: "Feb 2017"
output:
  pdf_document: default
  html_document: default
---


# Visualising phylogenies in R
The aims of this practical are to introduce you to some of the fun ways you can visualise phylogenetic trees in R.
The code and examples are based on Liam Revell's book chapter [here](http://faculty.umb.edu/liam.revell/pdfs/Revell_2014.MPCM-chapter.pdf).

Note that some of these plots won't work very well in RStudio. 
You may need to click the zoom button to see them, or resort to using the old R GUI instead.

__REMEMBER__

- Download all of the data for the practical into a folder somewhere on your computer.
- Set your working directory to this folder.
- Start a new script for this practical.
- Note that this will work better in the R console than RStudio

You will also need to install the following packages: 

* `ape`
* `phytools`

***

First load the packages we need and the data.

```{r, message = FALSE}
library(ape)
library(phytools)
```
Later we will use some Greater Antillean *Anolis* lizard data to add data to a phylogeny.
Before we can add data to our tree, we need to load the data. 

```{r}
anoledata <- read.csv("anole.data.csv", header = TRUE)

# Look at the data
str(anoledata)
```
Now let's visualise some phylogenies!

## Loading your phylogeny and data into R
### Reading in a phylogeny from a file
To load a tree you need either the function `read.tree` or `read.nexus`.
`read.tree` can deal with a number of different types of data (including DNA) whereas `read.nexus` reads NEXUS files. 
`elopomorph.tre` is not in NEXUS format so we use `read.tree`. 

```{r}
fishtree <- read.tree("elopomorph.tre")
```

### Reading in a phylogeny that is already built into R
The bird and anole phylogenies are already built into R so we don't need to read them in using `read.tree`.
Instead we just use:

```{r}
data(bird.orders)
data(anoletree)
```

## Basic tree viewing in R
We'll use the Elopomorpha (eels and similar fishes) tree to start as it is simple.

Let's examine the tree by typing:

```{r}
str(fishtree)
```

`fishtree` is a fully resolved tree with branch lengths. 
There are 62 species and 61 internal nodes. 
We can plot the tree by using the `plot.phylo` function of `ape`. 
Note that we can just use the function `plot` to do this as R knows if we ask it to plot a phylogeny to use `plot.phylo` instead!

```{r}
# Set the margins to make the tree visible
par(mar = c(0,0,0,0))
plot(fishtree, cex = 0.5)
```

`cex = 0.5` reduces the size of the tip labels so we can read them. 
We can also zoom into different sections of the tree that you're interested in:

```{r}
par(mar = c(0,0,0,0))
zoom(fishtree, grep("Gymnothorax", fishtree$tip.label), subtree = FALSE, cex = 0.8)
```

This just gives you the tree for *Micropterus* species but you can also see how the species fit into the rest of the tree using:

```{r}
par(mar = c(0,0,0,0))
zoom(fishtree, grep("Gymnothorax", fishtree$tip.label), subtree = TRUE, cex = 0.8)
```

Note that `zoom` automatically sets the plotting window to display two plots at once. 
To reset this to one plot only use:

```{r}
par(mfrow = c(1, 1))
```

To get further options for the plotting of phylogenies:

```{r, eval = FALSE}
?plot.phylo
```

Note that although you can use `plot` to plot the phylogeny, you need to specify `plot.phylo` to find out the options for plotting trees. You can change the style of the tree (`type`), the color of the branches and tips (`edge.color`, `tip.color`), and the size of the tip labels (`cex`). 
Here's an fun/hideous example! 

```{r}
par(mar = c(0,0,0,0))
plot(fishtree, type = "fan", edge.color = "deeppink", tip.color = "springgreen", 
	 cex = 0.5)
```

Or try

```{r}
par(mar = c(0,0,0,0))
plot(fishtree, type = "c", edge.color = "darkviolet", tip.color = "hotpink", 
	 cex = 0.5)
```

## Adding trait data to trees in R
*Note that the theory behind this is covered in more detail in the "Macroevolutionary Models in R" practicals part 1 and 2.*
### Ancestral state reconstructions on discrete data
We will use the bird data.
Remember we already loaded the phylogeny and data as follows:

```{r}
data(bird.orders)
```

First we will invent some data for each bird order that we can reconstruct along the tree. 
This creates three variables, 0, 1 and 2. 

```{r}
mydata <- factor(c(rep(0, 5), rep(1, 13), rep(2, 5)))
mydata
```

We can then use the `ape` function `ace` to reconstruct ancestral characters along the nodes of the tree. 
`type = d` means the character to be reconstructed is discrete.

```{r}
ancestors <- ace(mydata, bird.orders, type = "d")
```

Don't worry about the NAs produced here.
Now we can plot this on a phylogeny. 
First decide which colours we'd like.
To look at a list of colours in R type in `colors()`.

```{r}
colours <- c("cornflowerblue", "cyan4", "goldenrod")
```

Now plot the tree and add labels to the tips and the nodes using the results in `ancestors`. 
We use `label.offset = 1` to move the labels to the right a bit so the pie charts will fit...

```{r}
par(mar = c(0,0,0,0))
plot(bird.orders, label.offset = 1)
tiplabels(pch = 21, bg = colours[as.numeric(mydata)], cex = 2, adj = 1)
nodelabels(pie = ancestors$lik.anc, piecol = colours)
```

`pch = 21` sets the tip labels to be unfilled circles, `bg` defines the colours of the circles using the list of colours we provided, and ordering them based on what the species value was for mydata (i.e. 0, 1 or 2).
`cex = 2` doubles the point size, and `adj = 1` moves the tip labels sideways so they don't obscure the ends of the branches.
`pie` makes pie charts coloured using the ancestral state reconstructions in `ancestors`, and `piecol` tells it to use the colours we have defined.

### Ancestral state reconstructions on continuous data
We are going to use the *Anolis* data to create a phylogeny with different colours for different observed and reconstructed body sizes (snout-to-vent length, SVL). 
Remember we already loaded the phylogeny and data as follows:

```{r}
data(anoletree)
anoledata <- read.csv("anole.data.csv", header = TRUE, row.names = 1) 
```

Note the names in `anoledata` are the species names without the Genus. 
In the phylogeny the species names are `Anolis_species`. 
So to get the two to match we need to add `Anolis_` to each name.

```{r}
rownames(anoledata) <- paste("Anolis", rownames(anoledata), sep = "_")
```
`paste` just sticks together `Anolis` with the names in `anoles` already with an underscore (`_`) separating them (`sep = "_"`)

We then need to make sure the order of the species in the data matches that of the phylogeny.

```{r}
anoledata <- anoledata[anoletree$tip.label, ]
```

Next we make a matrix containing only the SVL values for each *Anolis* species:

```{r}
SVL <- as.matrix(anoledata)[,"AVG.SVL"]
```

This code selects only the variable `AVG.SVL` from `anoledata` (square brackets subset in R in the form [rows, columns]), and then uses `as.matrix` to make this into a matrix.

Take a look at `SVL`

```{r}
head(SVL)
```

We then use the function `contMap`. 
`contMap` creates a tree with a mapped continuous character, i.e. where the value of the character is known at the tips, and estimated along the tree. 
The estimating of the character along the tree uses a Maximum Likelihood estimation procedure.
Here we will tell `contMap` not to automatically plot the tree (using `plot = FALSE`) so we can make some modifications.

```{r}
SVLplot <- contMap(anoletree, SVL, plot = FALSE)
```
 For some reason this code breaks...

Finally let's plot the tree as a fan (`legend = 10` just spreads the legend out so it is readable).

```{r}
par(mar = c(0,0,0,0))
plot(SVLplot, type = "fan", legend = 10)
```

## Phylomorphospace plots in R
We are going to use the *Anolis* data to create a phylomorphospace plot. 
These plots show where species fall within a morphospace (usually based on principal components analyses of morphological variables), and then reconstructs values for the nodes of the phylogeny so the whole phylogeny can be projected into the morphospace.

We already loaded the phylogeny and data as follows:

```{r}
data(anoletree)
anoledata <- read.csv("anole.data.csv", header = TRUE, row.names = 1)
```

As we saw above, the names of `anoles` are the species names without the Genus. 
In the phylogeny the species names are `Anolis_species`. 
So to get the two to match we need to add `Anolis_` to each name.

```{r}
rownames(anoledata) <- paste("Anolis", rownames(anoledata), sep = "_")
```

`paste` just sticks together `Anolis` with the names in `anoles` already with an underscore (`_`) separating them (`sep = "_"`)

Next we again need to make sure the order of the species in the data matches that of the phylogeny.

```{r}
anoledata <- anoledata[anoletree$tip.label, ]
```

We can now run a phylogenetic principal components analysis on our morphological trait data and extract the PC scores for each PC axis using `$S`.

```{r}
PC <- phyl.pca(anoletree, anoledata)$S
```

To plot the morphospace with colours, we need to define them. 
Here we choose six colours, and matched them with the ecomorph types listed [here](https://en.wikipedia.org/wiki/Anolis_ecomorphs).
These were built into `anoletree`.


```{r}
colors <- setNames(c("cornflowerblue", "goldenrod", "chartreuse", "hotpink",
	                 "orangered", "darkviolet"), 
                     sort(unique(getStates(anoletree, "tips"))))
```

Finally we can make the plot, and colour by ecomorph. 
`label = "off"` suppresses the printing of tip labels which keeps things a bit tidier.

```{r}
phylomorphospace(anoletree, PC[,1:2], label = "off", node.by.map = TRUE, colors = colors)
```

## Phylogenies and maps!
We can also plot phylogenies attached to maps showing where species come from.
We will use the bird orders tree again as it's small.

```{r}
data(bird.orders)
```

We don't know where the birds come from, so we will simulate some latitude and longitude data as follows. 
This uses a function that simulates traits along the phylogeny, so close relatives should end up with more similar latitude and longitude values.
I've used high rates of evolution (`sig2`) to force the points to be spread out. 
Of course these values will be nonsensical for birds!

```{r}
lat <- fastBM(bird.orders, sig2 = 100, bounds = c(-90, 90))
long <- fastBM(bird.orders, sig2 = 150, bounds = c(-180, 180))
```

Then we use the function `phylo.to.map` to match the locations with the world map. This produces some output in the form `objective 98` etc. (I have suppressed it here). 

```{r, message = FALSE}
phylomap <- phylo.to.map(bird.orders, cbind(lat,long), plot = FALSE)
```

We can either put the phylogeny next to the map and draw lines to the places they come from using `type = "phylogram"`...

```{r}
par(mar = c(0,0,0,0))
plot(phylomap,type = "phylogram", asp = 1.3, mar = c(0.1, 0.5, 3.1, 0.1))
```

...or plot the phylogeny directly onto the map using `type = "direct"`

```{r}
plot(phylomap,type = "direct", asp = 1.3, mar = c(0.1, 0.5, 3.1, 0.1))
```

## A new package for visualising trees: ggtree
`ggtree` is a newly released package that works with the popular `ggplot` framework in R. 
`ggplot` works by adding layers of different features together. 
A standard `ggplot` includes a layer that assigns the x and y variables then can have lots of other layers of points, lines, shapes and now with `ggtree` phylogenies.

__Unfortunately this is broken in the most recent version of R. I've left this in, just in case they fix it soon__

To use `ggtree` you will have to install it, then load it using:

```{r, eval = FALSE}
library(ggtree)
```

Let's replicate some of the stuff we did above but using `ggtree`. 
As always choose whichever method you prefer when you're doing this with your own data.

```{r, eval = FALSE}
ggtree(fishtree) +
geom_text(aes(label=label), size = 1)
```

Note that the layers here are the tree, and then the tip labels. 
Layers are separated by `+`

We can also zoom in on sections like with `ape`.

```{r, eval = FALSE}
gzoom(fishtree, grep("Gymnothorax", fishtree$tip.label))
```

And we can have different shapes of phylogeny, different colours and even different line types.

```{r, eval = FALSE}
ggtree(fishtree, color = "steelblue")
ggtree(fishtree, layout = "circular")
ggtree(fishtree, linetype = "dotted")
```
     
`ggtree` is really great for highlighting sections of a tree.
First look at the nodes on your tree and pick an interesting one...

```{r, eval = FALSE}
ggtree(fishtree) + 
  geom_text(aes(label = node))
```

Then highlight and/or annotate...

```{r, eval = FALSE}
ggtree(fishtree) +
geom_hilight(node = 73, fill = "steelblue", alpha = 0.6) +
geom_hilight(node = 97, fill = "darkgreen", alpha = 0.6)
```

This package has a huge amount of promise, but still a few bugs and weird features (it's designed by geneticists who are used to dealing with short species labels, hence the issues getting our species names to actually fit on the plot). The help files are also currently a bit sparse and some of the vignette examples don't work. But if you're likely to need to do any of this kind of stuff I recommend checking out the vignette and the paper:

* Guangchuang Yu, David Smith, Huachen Zhu, Yi Guan, Tommy Tsan-Yuk Lam. ggtree: an R package for visualization and annotation of phylogenetic trees with their covariates and other associated data. _Methods in Ecology and Evolution_ 2016, doi:10.1111/2041-210X.12628.
