---
title: "Macroevolutionary models in R - Part 1 - continuous traits"
author: "Natalie Cooper (natalie.cooper@nhm.ac.uk)"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---


# Macroevolutionary models in R: Part 1 - continuous traits

The aims of this practical are to learn how to use R to fit macroevolutionary models in R to continuous traits.

We will be using the evolution of magical creature life-history variables as an example. 
The data includes body mass (average adult size at rest) in kg, social status (1 = solitary, 2 = social), habitat (1 = terrestrial, 2 = aquatic, 3 = volant) and magical power (in thaum - with thanks to Terry Pratchett for the units). 
These data are invented, so please don't get too upset if I've misclassified anything!

__REMEMBER__

- Download all of the data for the practical into a folder somewhere on your computer.
- Set your working directory to this folder (if using `.Rmd` this will happen automatically).
- Start a new script for this practical, or edit the `.Rmd` file after saving it with a new name.

You will also need to install the following packages: 

* `ape`
* `geiger`
* `OUwie`

This practical is in two parts, Part 2 deals with discrete traits.

***

## Preparing for the analysis
### Load packages, read in the data and the tree
This is the same as we did in the PGLS practical, so I won't give detailed instructions here.

```{r, message = FALSE}
# Load packages
library(ape)
library(geiger)
library(OUwie)
```

```{r, message = FALSE}
# Read in data
magicaldata <- read.csv("magicalcreatures.csv")
# Check data is loaded correctly
str(magicaldata)
```

```{r}
# Read in tree
magicaltree <- read.nexus("magicaltree.nex") 
# Check tree is loaded correctly
str(magicaltree)
```

### Modify the tree and data so they can be used in the analyses.
Again we did this in the PGLS practical. 
Please remind yourself of what these steps are needed for.

```{r}
# Ensure tree is fully bifurcating
magicaltree <- multi2di(magicaltree) 

# Replace spaces with underscores in the species names
magicaldata$Species <- gsub(" ", "_", magicaldata$Species)

# Add species names to row names
row.names(magicaldata) <- magicaldata$Species
```

For some weird reason the `geiger` function we need (`treedata` see below) won't work if you input a dataset with variables that are characters i.e. words or letters. 
Our taxonomic variable Species is a character so we need to exclude it from the data.
Note for your own data you'd need to remove all character variables (or recode them as 0,12 etc.).
We will do this by making a new dataset called magicaldata2.

```{r}
magicaldata2 <- magicaldata[, 2:5]
```

Here the [ ] tells R we want to subset the dataset. 
R data frames are always described by [X,Y] where X is rows and Y is columns. 
So [1, 1] will select the entry in the first column and the first row of the data frame. 
[, 2:2] selects all rows but only columns 2 to 5. 
These are the columns containing our numeric variables.

We then need to match the species in tree to those in the dataset as in the PGLS practical. 
__Note that we are using `magicaldata2` here.__

```{r, warning = FALSE}
match.species <- treedata(magicaltree, magicaldata2)

mytree <- match.species$phy
mydata <- match.species$data
```

***
## Models of continuous trait evolution
For fitting models of evolution to continuous data we will use the `fitContinuous` function in the R package `geiger`.
`fitContinuous` is a likelihood based method, so the output will give the maximum likelihood (ML) estimates of the parameters. 
Bayesian methods are becoming preferred for these kinds of analyses and `fitContinuousMCMC` will perform these analyses. 
However, due to time constraints we will not cover this function.
As an example, let's look at the evolution of `log(body size)` in magical creatures. 
We'll fit three evolutionary models – the Brownian motion (BM) model, the Ornstein-Uhlenbeck (OU) model and the Early Burst (EB) model. 
`fitContinuous` can also fit several other models. 
For more details look at the help file by typing: `?fitContinuous`

### The Brownian motion (BM model)
The Brownian motion model (Cavalli-Sforza and Edwards, 1967; Felsenstein, 1973) is assumed to be the underlying mode of evolution in the majority of phylogenetic comparative methods (though this assumption is rarely tested; Freckleton and Harvey 2006). 
In the model, a trait $X$ evolves at random at a rate $\sigma$:

$dX(t) = \sigma dW(t)$

where $W(t)$ is a white noise function and is a random variate drawn from a normal distribution with mean 0 and variance $\sigma^2$. This model assumes that there is no overall drift in the direction of evolution (hence the expectation of $W(t)$ is zero) and that the rate of evolution is constant.
Because the direction of change in trait values at each step is random, Brownian motion is often described as a “random walk” (note that you can also fit models where $W(t)$ is not zero and there is drift in the direction of evolution. 
This is known as the drift model, or Brownian motion with a trend. We will not cover this here).

The model assumes the correlation structure among trait values is proportional to the extent of shared ancestry for pairs of species. 
This means that close relatives will be more similar in their trait values than more distant relatives. 
It also means that variance in the trait will increase (linearly) in proportion to time.
The model has two parameters, the Brownian rate parameter, $\sigma^2$ and the state of the root at time zero, $X(0)$. 
`fitContinuous` estimates $\sigma^2$ and $X(0)$. 
Note that $\sigma^2$ can be used as a measure of the rate of trait evolution (Cooper and Purvis, 2010; Cooper et al., 2011).

### The Ornstein-Uhlenbeck (OU) model
The Ornstein-Uhlenbeck (OU) model (Hansen, 1997; Butler and King, 2004) is a random walk where trait values are pulled back towards some “optimal” value with an attraction strength proportional to the parameter $\alpha$. 
The model has the following form:

$dX(t) = -\alpha(X(t) - \mu) + \sigma dW(t)$

Note that this model has two parameters in addition to those of the Brownian model, $\alpha$ and $\mu$.
The parameter m is a long-term mean, and it is assumed that species evolve around this value. 
$\alpha$ is the strength of evolutionary force that returns traits back towards the long-term mean if they evolve away from it. 
$\alpha$ is sometimes referred to as the “rubber band” parameter because of the way it forces traits back towards $\mu$.
The OU model was introduced to population genetics by Lande (1976) to model stabilizing selection in which the mean was recast as a fitness optimum on an adaptive landscape. 
The process operating in comparative data is analogous, although clearly is not stabilizing selection (despite being sometimes referred to as such).
The model has four parameters, the Brownian rate parameter, $\sigma^2$, the state of the root at time zero, $X(0)$, the attraction strength or “rubber band” parameter, $\alpha$, and the long-term mean, $\mu$.
`fitContinuous` estimates $\sigma^2$, $X(0)$, and $\alpha$. 
It does not estimate $\mu$ but in this implementation of the model, $\mu$ is equivalent to $X(0)$. 
Note that if $\alpha$ is close to zero then evolution is approximately Brownian.

### The Early Burst (EB) model
The Early Burst (EB) model (Harmon et al. 2010, also called the ACDC model; Blomberg et al. 2003) is a Brownian motion/random walk model where the rate of evolution decreases exponentially through time under the model:

$r(t) = \sigma^2e(at)$

Where $r(t)$ is the rate of evolution at time $t$, $\sigma^2$ is the initial value of the Brownian rate parameter, i.e. the initial rate of evolution, $a$ is the rate change parameter, and $t$ is time. 
The value of $a$ is generally less than or equal to 0 (note that you can force $a$ to be greater than zero by changing the bounds - see section 3.4 - however, this will only work if you have fossil species in your data; Slater et al. 2012). 
When a is negative, rates of evolution decrease through time. 
The model fits traits where diversification occurs most rapidly early in a lineage and slows as the
lineage approaches the present, so that subclades tend to retain their differences through time. 
This is consistent with a clade radiating adaptively into a fixed set of niches and has been used as evidence of niche-filling modes of evolution (Harmon et al., 2010; Cooper and Purvis, 2010).
The model has three parameters, the Brownian rate parameter, $\sigma^2$, the state of the root at time zero, $X(0)$, and the rate of change parameter, $a$. 
`fitContinuous` estimates $\sigma^2$, $X(0)$ and $a$. 
Note that if $a$ is close to zero then evolution is approximately Brownian.
Note that although many people report a values when reporting the results of fitting an Early Burst model, it is often more intuitive to report the rate half-life, $t_{\frac{1}{2}}$ (Slater and Pennell, 2014). 
This is calculated as:

$t_{\frac{1}{2}} = \frac{log(2)}{|a|}$

It can be interpreted as the time it takes for the rate of evolution of the trait to halve.

## Fitting the models using `fitContinuous`
First we will fit a Brownian motion (BM) model to log(body mass):

```{r}
BM <- fitContinuous(mytree, log(mydata[,"BodySize_kg"]), model = c("BM"))
```

To look at the output type:

```{r}
BM
```

The maximum likelihood estimates (`lnL`) of the model parameters are found near the top of the output. 
In a Brownian motion (BM) model we estimate the Brownian rate parameter, $\sigma^2$ or `sigsq` in the output above, which is `2.208` and the value of the trait at the root of the tree, $X(0)$ or `z0` in the output above, which is `4.304`.
Other useful things in the output are the maximum-likelihood estimate (`lnL`) of the model (log-likelihood), the Akaike Information Criterion (`AIC`), sample-size corrected AIC (`AICc`) and the number of model parameters (free parameters) also known as $k$ in the literature. 
We will return to the AIC values below.

To fit an Ornstein-Uhlenbeck model to log(body mass) we only need to change the model in the formula we used above:

```{r, message = FALSE}
OU <- fitContinuous(mytree, log(mydata[,"BodySize_kg"]), model = c("OU"))
```

To look at the output type:

```{r}
OU
```

As above, the maximum likelihood estimates (`lnL`) of the model parameters are found near the top of the output. 
In an Ornstein-Uhlenbeck (OU) model we estimate the Brownian rate parameter, $\sigma^2$ or `sigsq` in the output above, the value of the trait at the root of the tree, $X(0)$ or `z0` in the output above, and the selection strength parameter, $\alpha$ or `alpha` in the output above. 
As `alpha = 2.718` here, it suggests that there is evolution towards a body mass optimum.

Finally, to fit an Early Burst (EB) model to log(body mass):

```{r}
EB <- fitContinuous(mytree, log(mydata[,"BodySize_kg"]), model = c("EB"))
```

To look at the output type:

```{r}
EB
```

As above, the maximum likelihood estimates (`lnL`) of the model parameters are found near the
top of the output. In an Early Burst (EB) model we estimate the Brownian rate parameter, $\sigma^2$ or
`sigsq` in the output above, the value of the trait at the root of the tree, $X(0)$ or `z0` in the output
above, and the rate of change parameter, $a$. Here `a` is very close to 0 indicating that the rate of log(body
mass) evolution in magical creatures has not decreased through time.

We can also extract the rate half-life, $t_{\frac{1}{2}}$, for this model as follows:

```{r}
log(2)/abs(EB$opt$a)
```
For these data, the rate half-life is almost 700,000 time units, much greater than the total time represented on the tree!
This means over the course of magical creature evolution, body size rate has not halved (yet). 

### Comparing models of evolution
Often we want to know which of the models fits our variable best. 
We can use `fitContinuous` to fit the models we are interested in and can then compare them using AIC. 
We can extract the AICs from the models we fitted above as follows:

```{r}
BM$opt$aic
```

```{r}
OU$opt$aic
```

```{r}
EB$opt$aic
```

The “best” model is the one with the smallest AIC, in this case the OU model. 
There is much debate about how big of a difference in AIC values can be classed as substantial improvement to a model fit (it usually ranges from 2-10 AIC units). 
Generally we use 4 units, so EB probably doesn't fit substantially better than BM, but OU is substantially better than BM and EB.

Alternatively we can use $\Delta$AIC or AIC weights to compare our models using the following code and the `geiger` function `aicw`:

```{r}
aic.scores <- setNames(c(BM$opt$aic, OU$opt$aic, EB$opt$aic), c("BM","OU","EB"))
aicw(aic.scores)
```

`aicw` outputs the AIC (`fit`), $\Delta$AIC (`delta`) and AIC weights (`w`) for each of the models we fitted.
The best model is the model with $\Delta$AIC = 0 or with AICw closest to 1. 
Using $\Delta$AIC we can conclude that the OU model is the best fit to the data.

### Problems with convergence and bounds
Above we have mentioned the default bounds on each parameter. 
Sometimes these need to be changed because the model will not converge. 
This happens when the likelihood surface has long flat ridges that cause the likelihood search to get “stuck” (this is particularly common under the OU model). 
You can change bounds with the bounds argument in `fitContinuous`. 
Several bounds can be given at a time e.g. `bounds = list(sigsq = c(0, 0.1), alpha = c(0, 1))` would constrain both
the $\sigma^2$ and $\alpha$ parameters.

For example, if an OU model keeps getting stuck you could try changing the lower bound on $\alpha$:

```{r, message = FALSE}
OU <- fitContinuous(phy = mytree, dat = log(mydata[,"BodySize_kg"]), model = c("OU"), 
                    bounds = list(alpha = c(0.01, 1)))
```

This example gives a warning message because `alpha` is over 1, so when we make the upper bound smaller the method ends up giving us this value instead because it's as close to 1 as we are allowing the model to go.

***

## Practice Questions

__Remember to save your code and answers for the assessment__

See part 2 of the practical.

***

## References

* Blomberg, S. P., T. Garland, and A. R. Ives. 2003. Testing for phylogenetic signal in comparative data: behavioral traits are more labile. Evolution 57:717–745.
* Butler, M. A. and A. A. King. 2004. Phylogenetic comparative analysis: a modeling approach for adaptive evolution. The American Naturalist 164:683–695.
* Cavalli-Sforza, L. L. and A. W. Edwards. 1967. Phylogenetic analysis. models and estimation procedures. American Journal of Human Genetics 19:233.
* Cooper, N., R. P. Freckleton, and W. Jetz. 2011. Phylogenetic conservatism of environmental niches in mammals. Proceedings of the Royal Society B: Biological Sciences 278:2384–2391.
* Cooper, N. and A. Purvis. 2010. Body size evolution in mammals: complexity in tempo and mode.The American Naturalist 175:727–738.
* Cooper, N., Thomas, G.H., Venditti, C., Meade, A. & Freckleton, R.P. (2016b) A cautionary note on the use of ornstein-uhlenbeck models in macroevolutionary studies. Biological Journal of the Linnaean Society
* Felsenstein, J. 1973. Maximum likelihood and minimum-steps methods for estimating evolutionary trees from data on discrete characters. Systematic Biology 22:240–249.
* Freckleton, R. P. and P. H. Harvey. 2006. Detecting non-brownian trait evolution in adaptive radiations. PLoS Biology 4:e373.
* Hansen, T. F. 1997. Stabilizing selection and the comparative analysis of adaptation. Evolution Pages 1341–1351.
* Harmon, L. J., J. B. Losos, T. Jonathan Davies, R. G. Gillespie, J. L. Gittleman, W. Bryan Jennings, K. H. Kozak, M. A. McPeek, F. Moreno-Roark, T. J. Near, et al. 2010. Early bursts of body size and shape evolution are rare in comparative data. Evolution 64:2385–2396.
* Lande, R. 1976. Natural selection and random genetic drift in phenotypic evolution. Evolution 30:314–334.
* Slater, G. J., L. J. Harmon, and M. E. Alfaro. 2012. Integrating fossils with molecular phylogenies improves inference of trait evolution. Evolution 66:3931–3944.
* Slater, G. J. and M.W. Pennell. 2014. Robust regression and posterior predictive simulation increase power to detect early bursts of trait evolution. Systematic Biology 63:293–308.
